---
title: "Creating a workflow step"
output:
  "BiocStyle::html_document":
    dev: png
    df_print: "kable"
vignette: >
  %\VignetteIndexEntry{Creating a workflow step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(cowplot)
library(GCIMS)
```

# WARNING

This vignette still uses some internals of the GCIMS package. We will be exporting
helper functions and changing the design of the objects to make it more robust,
so if you plan to create custom methods outside the GCIMS package, please
stay in touch and coordinate with us.


# Introduction

In this vignette we show how samples can be modified in a batch, using delayed
operations. GCIMS uses delayed evaluations were possible, so dataset modifications can
be executed in a more efficient way and without using too much RAM.

You may want to read this if you want to add support for a new algorithm that
modifies all the samples in a dataset.

```{r}
# The folder where we will download the samples:
samples_directory <- "threeketones"

# Download the ketones dataset:
tryCatch({
  download_three_ketones_dataset(samples_directory)
  message("Download successful")
}, error = function(e) {
  message("The download of the samples did not succeed. The vignette can't continue.")
  message(conditionMessage(e))
  knitr::knit_exit()
})

```


# Batch modifying all samples

You will first have to define a function that takes one sample
as an argument and returns that sample modified:

```{r}
cutDrift_Sample <- function(x, dt_from, dt_to) {
  # dt_from and dt_to are indices of the drift time vector.
  # This function crops a sample in the drift time axis.
  
  # Check that we are given a sample
  if (!inherits(x, "GCIMSSample")) {
    stop("Internal error, this will never happen")
  }
  
  # In the future we will provide accessors instead of using @ slots directly.
  x@data <- x@data[dt_from:dt_to,]
  x@drift_time <- x@drift_time[dt_from:dt_to]
  
  # Check that we return a valid sample:
  if (!validObject(x)) {
    stop("The sample is not valid.")
  }
  return(x)
}
```

Then you need a function that applies to the whole dataset and returns the dataset.
You can use this template for now:

```{r}
cutDrift_Dataset <- function(x, dt_from, dt_to) {
    delayed_op <- GCIMS:::GCIMSDelayedOp(
      name = "cutDrift", # Give a human-readable name
      fun = cutDrift_Sample, # The function that is applied to the sample
      params = list(dt_from = dt_from, dt_to = dt_to) # The arguments that function needs
    )
    x <- GCIMS:::appendDelayedOp(x, delayed_op)
    # Return the dataset, invisibly:
    invisible(x)
}
```


## Example:

Load the dataset:

```{r}
annotations <- read.csv(file.path(samples_directory, "annotations.csv"))
dataset <- GCIMSDataset(
  annotations,
  base_dir = samples_directory,
  on_ram = TRUE # You probably should set this to FALSE if you have more 
                # than a handful of samples. See ?GCIMSDataset.
)
dataset

```

Get the first sample of the dataset, and see the drift time vector at indices 500 and 1200:

```{r}
sample_before <- getSample(dataset, 1)
dtime_before <- dtime(sample_before)
```

```{r}
dtime_before[1000]
dtime_before[1500]
```

Cut the sample:

```{r}
dataset <- cutDrift_Dataset(dataset, dt_from = 1000, dt_to = 1500) # indices
dataset
```

The dataset has now a pending operation, we can execute it:

```{r}
realize(dataset)
```

And get the sample, modified:

```{r}
sample_after <- getSample(dataset, 1)
dtime_after <- dtime(sample_after)
```


```{r}
dtime_after[1]
dtime_after[length(dtime_after)]
```

Same sample before and after:

```{r fig.height=14, fig.width=7}
plot_grid(
  plot(sample_before) + labs(title = "Before"),
  plot(sample_after) + labs(title = "After"),
  nrow = 2
)
```


# Save additional information within each sample

Modifying your samples is something nice. Sometimes you want to save some
additional information in the sample that can help you analyze the results.

You can do that with the `proc_params` slot.

```{r}
cutDrift2_Sample <- function(x, dt_from, dt_to) {
  # dt_from and dt_to are indices of the drift time vector.
  # This function crops a sample in the drift time axis.
  
  # Check that we are given a sample
  if (!inherits(x, "GCIMSSample")) {
    stop("Internal error, this will never happen")
  }
  
  # Get the original range of the drift time
  original_dt_length <- length(dtime(x))
  

  # In the future we will provide accessors instead of using @ slots directly.
  x@data <- x@data[dt_from:dt_to,]
  x@drift_time <- x@drift_time[dt_from:dt_to]
  
  new_dt_length <- length(dtime(x))
  
  # Save it in the proc_params list:
  x@proc_params$cutDrift <- list(
    original_dt_length = original_dt_length,
    new_dt_length = new_dt_length
  )
  
  # Check that we return a valid sample:
  if (!validObject(x)) {
    stop("The sample is not valid.")
  }
  return(x)
}
```

We use the same function for the dataset:

```{r}
cutDrift2_Dataset <- function(x, dt_from, dt_to) {
    delayed_op <- GCIMS:::GCIMSDelayedOp(
      name = "cutDrift2", # Give a human-readable name
      fun = cutDrift2_Sample, # The function that is applied to the sample
      params = list(dt_from = dt_from, dt_to = dt_to) # The arguments that function needs
    )
    x <- GCIMS:::appendDelayedOp(x, delayed_op)
    # Return the dataset, invisibly:
    invisible(x)
}
```


## Example

```{r}
dataset <- GCIMSDataset(
  annotations,
  base_dir = samples_directory,
  on_ram = TRUE # You probably should set this to FALSE if you have more 
                # than a handful of samples. See ?GCIMSDataset.
)
dataset

```

```{r}
dataset <- cutDrift2_Dataset(dataset, dt_from = 1000, dt_to = 1500) # indices
dataset
```

The dataset has now a pending operation, we can execute it:

```{r}
realize(dataset)
```

And get the sample, modified:

```{r}
sample_after <- getSample(dataset, 1)
```

We have those parameters saved:

```{r}
sample_after@proc_params$cutDrift
```



# Extracting and aggregating information from samples into the dataset

In the previous example, we have to manually `getSample()` on each sample
to get those additional results. This is going to be very slow.

What we would like to have is a data frame with three columns: the sample name, the original dt_length
and the new dt_length, so we have a summary for all the dataset.

For this to happen fast and efficiently, we need to do it in two steps:

- Extraction: Extract (a small amount of) information from a sample.
- Aggregation: Aggregate the extracted information (e.g. to build your dataframe)

## Extraction

The extraction is usually a simple function that extracts from a sample
the information you need. Some workflow steps use it to extract normalization
factors, other workflow steps use it to extract the RIC or the TIS.

Avoid extracting the whole data matrix, since it will use too much RAM.

The function must take a GCIMSSample, and may return whatever you need. If
you need many things, return a list with all of them.

```{r}
cutDrift_ExtractSample <- function(x) {
  # Check that we are given a sample
  if (!inherits(x, "GCIMSSample")) {
    stop("Internal error, this will never happen")
  }

  # Extract only what you need:
  extracted_object <- x@proc_params$cutDrift
  return(extracted_object)
}
```

### Check that it works:

You can check that your `*_ExtractSample` works by applying it to one sample, if it
has been processed:

```{r}
cutDrift_ExtractSample(sample_after)
```


## Aggregation

The `realize()` function will extract the information from all your samples and
save all the extracted information on a list, as long as the number of samples
in your dataset.

Then, it will call the aggregation function, using the `GCIMSDataset` as first
argument and the list of extracted information in the second argument.

We must write the aggregation function to convert our objects into whatever
we want (a data frame in our example) and save the outcome in the dataset object.

```{r}
cutDrift_Aggregate <- function(x, extracted_objects) {
  # Check that we are given a dataset
  if (!inherits(x, "GCIMSDataset")) {
    stop("Internal error, this will never happen")
  }
  
  # Check that we have as many extracted objects as samples:
  if (length(extracted_objects) != length(sampleNames(x))) {
    # This should never happen:
    stop("Internal error, we always have an extracted object for each sample")
  }
  
  # Build the data frame that we want to generate
  df <- data.frame(
    SampleName = sampleNames(x),
    orig_dt_length = NA_integer_,
    new_dt_length = NA_integer_
  )
  # Fill it with information from the extracted objects:
  for (i in seq_along(extracted_objects)) {
    extracted_object <- extracted_objects[[i]]
    df$orig_dt_length[i] <- extracted_object$original_dt_length
    df$new_dt_length[i] <- extracted_object$new_dt_length
  }
  
  # Save our dataframe in the dataset object:
  x@envir$cutDrift <- df
  # We always return the whole dataset
  return(x)
}
```

We now pass those extraction and aggregation functions as well:

```{r}
cutDrift3_Dataset <- function(x, dt_from, dt_to) {
    delayed_op <- GCIMS:::GCIMSDelayedOp(
      name = "cutDrift3", # Give a human-readable name
      fun = cutDrift2_Sample, # The function that is applied to the sample
      params = list(dt_from = dt_from, dt_to = dt_to), # The arguments that function needs
      fun_extract = cutDrift_ExtractSample,
      fun_aggregate = cutDrift_Aggregate
    )
    x <- GCIMS:::appendDelayedOp(x, delayed_op)
    # Return the dataset, invisibly:
    invisible(x)
}
```


## Example

```{r}
dataset <- GCIMSDataset(
  annotations,
  base_dir = samples_directory,
  on_ram = TRUE # You probably should set this to FALSE if you have more 
                # than a handful of samples. See ?GCIMSDataset.
)
dataset

```

```{r}
dataset <- cutDrift3_Dataset(dataset, dt_from = 1000, dt_to = 1500) # indices
dataset
```

The dataset has now a pending operation, we can execute it:

```{r}
realize(dataset)
```

And then get our data frame:

```{r}
dataset@envir$cutDrift
```


Done!

# Session info


```{r}
sessionInfo()
```

