---
title: "Importing custom data formats"
output:
  "BiocStyle::html_document":
    dev: png
  "BiocStyle::pdf_document":
    latex_engine: lualatex
    df_print: "kable"
    dev: png
package: GCIMS
author: "Sergio Oller"
date: "`r format(Sys.Date(), '%F')`"
abstract: >
  This vignette shows how to import your custom data so it can be used
  with the GCIMS package. Data formats are typically vendor-dependant, and
  exports to CSV can have subtle differences.
vignette: >
  %\VignetteIndexEntry{Importing custom data formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

```{r setup}
library(GCIMS)
library(cowplot)
```


# Introduction

This vignette aims to show you how to create a GCIMSDataset object from your
own files, if those are not supported natively by the GCIMS package.

We do so, by showing how we can add support for importing CSV files.

The first step is to read the drift time, the retention time and the intensity
matrices from your data file. Then we create a GCIMSSample object.

Once we have solved that, we wrap all our written code into a function, and we
create the dataset.

# Creating GCIMSSample objects

To create a GCIMSSample object you need to have at least:

- A numeric vector with drift times in ms
- A numeric vector with retention times in s
- An intensity matrix, with dimensions $length(drift\_time) \times length(retention\_time)$

If your time vectors have different units, GCIMS will work, although you may see wrong labels in plots.
We plan to include support for more units in the future.

Let's imagine your sample is on a CSV file, with retention times on the first column,
drift times on the first row, and the corresponding intensity values.

We will now create two samples: sample1.csv and sample2.csv

```{r}
your_csv_file <- (
",0.0,0.1,0.2,0.3,0.4
0.0,  0, 20, 80, 84, 23
0.8,123,200,190,295, 17
1.6,230,300,200, 92, 15
2.4,120,150,120, 33, 22
3.2, 70,121, 74, 31, 34
")
write(your_csv_file, "sample1.csv")
write(your_csv_file, "sample2.csv")
```

You can read it using `read.csv()` or the `readr::read_csv()` function from the `readr` package.

```{r}
your_csv_file <- "sample1.csv"
csv_data <- read.csv(your_csv_file, check.names = FALSE)
```

Once loaded, your data will look like:

```{r}
csv_data
```

- The first column contains the retention time.
- The column names (with the exception of the first column, which is empty) contain the drift time
- The values of all columns but the first are the intensities

```{r}
retention_time <- csv_data[[1]]
drift_time <- as.numeric(colnames(csv_data)[-1])
intensity <- as.matrix(csv_data[,-1])
rownames(intensity) <- retention_time

```

The retention time:

```{r}
retention_time
```

The drift time:

```{r}
drift_time
```


The intensity matrix:

```{r}
intensity
```

With these three elements, we can create a GCIMSSample:

```{r}
s1 <- GCIMSSample(
  drift_time = drift_time, 
  retention_time = retention_time,
  data = intensity
)
s1
```


We are now ready to define a `parser` function that returns a GCIMSSample given a filename:

```{r}
GCIMSSample_from_csv <- function(filename) {
  csv_data <- read.csv(your_csv_file, check.names = FALSE)
  retention_time <- csv_data[[1]]
  drift_time <- as.numeric(colnames(csv_data)[-1])
  intensity <- as.matrix(csv_data[,-1])
  rownames(intensity) <- retention_time
  return(
    GCIMSSample(
      drift_time = drift_time,
      retention_time = retention_time,
      data = intensity
    )
  )
}
```

Try it with a single sample:

```{r}
s1 <- GCIMSSample_from_csv("sample1.csv")
s1
```


You can check the intensity matrix and you can plot the sample to check that
it behaves as expected:

```{r}
intensity(s1)
```


```{r}
plot(s1)
```


# Create the GCIMSDataset

Once you are satisfied with your function, prepare the phenotype data frame:

```{r}
pdata <- data.frame(
  SampleID = c("Sample1", "Sample2"),
  FileName = c("sample1.csv", "sample2.csv"),
  Sex = c("female", "male")
)
pdata
```

And create the dataset object, passing your `parser` function:

```{r}
ds <- GCIMSDataset$new(
  pData = pdata,
  base_dir = ".",
  parser = GCIMSSample_from_csv,
  scratch_dir = "GCIMSDataset_demo1"
)
ds
```

```{r}
cowplot::plot_grid(
  plot(getSample(ds, "Sample1")),
  plot(getSample(ds, "Sample2")),
  ncol = 2
)
```


You now have a dataset ready to be used.

# Session info

```{r}
sessionInfo()
```


