---
title: "Introduction to GCIMS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-gcims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GCIMS)
```

The GCIMS package allows you to import your Gas Chromatography - Ion Mobility Spectrometry samples,
preprocess them, align them one to each other and build a peak table with the relevant features.

This vignette will use a small dataset consisting of a mixture of FIXME ketones. The
dataset is included with the package

## Import data

Please start by preparing an Excel spreadsheet (or a CSV/TSV file if you prefer)
with your samples and their annotations. Please name the first column `SampleID`
and the second column `Filename`.

```{r}
## We will directly define the data frame here:
sample_annotations <- tibble::tribble(
  ~SampleID, ~Filename,
  "Ketone1", "ketone1.mea",
  "Ketone2", "ketone2.mea",
  "Ketone3", "ketone3.mea",
)
## But you can use the following code:
## Use this for excel spreadsheets:
# sample_annotations <- readxl::read_excel("your_excel_file.xlsx")
## Use this for CSV files:
# sample_annotations <- readr::read_csv("your_csv_file.csv")

sample_annotations
```

Raw data:

- Reading samples: gcims_read_samples
- Reading metadata: gcims_read_metadata
- Reshape sample: reshape_samples
- Create reference: gcims_create_reference
- Cut samples: gcims_cut_samples

Pre-Processing in Drift Time:

- Interpolating drift time: gcims_interpolate
- Drift time baseline removal: gcims_baseline_removal
- Drift time alignment: gcims_alignment

Pre-Processing in Retention Time:

- Interpolating retention time: gcims_interpolate
- Retention time baseline removal: gcims_baseline_removal
- Retention time alignment: gcims_alignment


Aligned data:

- ROIs detection à gcims_rois_selection
