---
title: "Introduction to GCIMS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction-to-gcims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GCIMS)
```

The GCIMS package allows you to import your Gas Chromatography - Ion Mobility Spectrometry samples,
preprocess them, align them one to each other and build a peak table with the relevant features.

This vignette will use a small dataset consisting of a mixture of FIXME ketones. The
dataset is included with the package

## Create work directories

The first steps, is to create all teh directories and folders where the samples are located and will be stored along all the process. 
The current working directory should be the folder where the raw data samples are located. Folders for each pre-processing step will be 
created.

```{r}
wd <- getwd()
data <- file.path(wd,'data')
reshaped <- file.path(wd,'reshaped')
cut <- file.path(wd,'cut')
interpd <- file.path(wd,'interpd')
filtd <- file.path(wd,'filtd')
bslnd <- file.path(wd,'bslnd')
alignd <- file.path(wd,'alignd')
interpr <- file.path(wd,'interpr')
filtr <- file.path(wd,'filtr')
bslnr <- file.path(wd,'bslnr')
alignr <- file.path(wd,'alignr')
```

## Import data

Please start by preparing an Excel spreadsheet (or a CSV/TSV file if you prefer)
with your samples and their annotations. Please name the first column `SampleID`
and the second column `Filename`.

```{r}
## We will directly define the data frame here:
sample_annotations <- tibble::tribble(
  ~SampleID, ~Filename,
  "Ketone1", "ketone1.mea",
  "Ketone2", "ketone2.mea",
  "Ketone3", "ketone3.mea",
)
## But you can use the following code:
## Use this for excel spreadsheets:
# sample_annotations <- readxl::read_excel("your_excel_file.xlsx")
## Use this for CSV files:
# sample_annotations <- readr::read_csv("your_csv_file.csv")

sample_annotations
```

Raw data:

- Reading samples: gcims_read_samples
All the samples for the working directory are going to be read and convert into the format needed for working with this package

```{r}
# READING SAMPLES
dir_in <- getwd()
dir_out <- data
gcims_read_samples(dir_in, dir_out, sftwr = 1)
```

- Reading metadata: gcims_read_metadata
An excel file containing all the metadata information needed is going to be read and all the metadata information will be included 
in the files

```{r}
# READING METADATA
dir_in <- data
samples <- 1:24
file <- "Metadata.xlsx"
gcims_read_metadata(dir_in, samples, file)
```

- Reshape sample: reshape_samples
An adjustment of the dimensions of the samples is performed. There are occasions where the samples do not measure exactly the same,
making very complicated working with them.

```{r}
# RESHAPE SAMPLES
dir_in <- data
dir_out <- reshaped
samples <- 1:24
gcims_reshape_samples(dir_in, dir_out, samples)
```

- Create reference: gcims_create_reference
A reference sample is created. This reference sample can be one selected sample, or just the mean of all the samples.

```{r}
# CREATE REFERENCE
dir_in <- reshaped
dir_out <- reshaped
samples <- 1:24
ref_number <- NULL
gcims_create_reference (dir_in, dir_out, samples, ref_number)
```

- Cut samples: gcims_cut_samples
The samples are cut for a better data manipulation. There are part of the data matrix that does not contain any information.
The sample cut can be performed at the beginning, but also at the end after all the pre-processing.

```{r}
#SAMPLES CUT
dir_in <- reshaped
dir_out <- cut
samples <- 1:24
rt_range <- c(0, 850)
dt_range <- c(7, 14)
gcims_cut_samples(dir_in, dir_out, samples, rt_range, dt_range)
```

Pre-Processing in Drift Time:

- Interpolating drift time: gcims_interpolate

```{r}
# INTERPOLATING DRIFT TIME
dir_in <- cut
dir_out <- interpd
samples <- 1:24
time <- "Drift"
gcims_interpolate(dir_in, dir_out, samples, time)
```

- Smoothing drift time: gcims_smoothing

```{r}
# DRIFT TIME FILTERING
dir_in <- interpd
dir_out <- filtd
time <- "Drift"
samples <- 1:24
gcims_smoothing(dir_in, dir_out, samples, time)
```

- Drift time baseline removal: gcims_baseline_removal

```{r}
# DRIFT TIME BASELINE REMOVAL
dir_in <- filtd
dir_out <- bslnd
time <- "Drift"
samples <- 1:24
gcims_baseline_removal(dir_in, dir_out, samples, time)
```

- Drift time alignment: gcims_alignment
COW Algorithm

```{r}
#ALIGNMENT DRIFT TIME
dir_in <- bslnd
dir_out <- alignd
samples <- 1:24
time <- "Drift"
seg_vector <- seq(from = 10, to = 30, by = 5)
slack_vector <- seq(from = 1, to = (seg_vector[length(seg_vector)] - 4), by = 3)
gcims_alignment(dir_in, dir_out, samples, time, seg_vector, slack_vector)
```


Pre-Processing in Retention Time:

- Interpolating retention time: gcims_interpolate

```{r}
# INTERPOLATING RETENTION TIME
dir_in <- alignd
dir_out <- interpr
samples <- 1:24
time <- "Retention"
gcims_interpolate(dir_in, dir_out, samples, time)
```

- Smoothing retention time: gcims_smoothing

```{r}
# RETENTION TIME FILTERING
dir_in <- interpr
dir_out <- filtr
time <- "Retention"
samples <- 1:24
gcims_smoothing(dir_in, dir_out, samples, time)
```

- Retention time baseline removal: gcims_baseline_removal

```{r}
# RETENTION TIME BASELINE REMOVAL
dir_in <- filtr
dir_out <- bslnr
time <- "Retention"
samples <- 1:24 
gcims_baseline_removal(dir_in, dir_out, samples, time)
```

- Retention time alignment: gcims_alignment
PTW algorithm

```{r}
#ALIGNMENT RETENTION TIME
dir_in <- bslnr
dir_out <- alignr
samples <- 1:24
time <- "Retention"
gcims_alignment(dir_in, dir_out, samples, time)
```


Data Visualization

- Spectra visualziation: gcims_plot_spec

```{r}
# PLOT SPECTRA
sample_num <- 1:24
rt_value <- NULL
dt_range <- NULL
dir_in <- alignr
gcims_plot_spec(dir_in, sample_num, rt_value, dt_range, colorby = "Name")
```

- Chromatograms visualization: gcims_plot_chrom

```{r}
# PLOT CHROMATOGRAMS
sample_num <- 1:24
dir_in <- alignr                          
dt_value <- NULL
rt_range <- NULL
gcims_plot_chrom(dir_in, sample_num, dt_value, rt_range, colorby = "Name")
```

- Sample image visualization: gcims_view_sample

```{r}
sample_num <- 1
dir_in <- alignr                          
dt_value <- NULL
rt_range <- NULL
gcims_view_sample(dir_in, sample_num, rt_value, dt_range)
```

Aligned data:

- ROIs detection à gcims_rois_selection
